---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
  keep_md: true
---

## Loading and preprocessing the data
```{r,echo=TRUE}
    library(dplyr)
    datadest <- "activity.zip"
    dataname <- unzip(datadest, list=TRUE)$Name
    unzip(datadest)
    dforig <- read.csv(dataname)
    df <- dforig[!is.na(dforig$steps),]
```

## What is mean total number of steps taken per day?
```{r}
stepsums <- as.data.frame(df %>% group_by(date) %>% summarise(totalsteps=sum(steps), .groups='drop'))
meansteps <- mean(stepsums$totalsteps)
medsteps <- median(stepsums$totalsteps)
hist(stepsums$totalsteps,xlab="Steps Per Day",main="Histogram of Daily Step Count")
```  
  
The mean number of steps taken per day is `r meansteps`.  The median number of steps taken per day is `r medsteps`.  

## What is the average daily activity pattern?
```{r}
stepmeans <- as.data.frame(df %>% group_by(interval) %>% summarise(meansteps=mean(steps), .groups='drop'))
maxstepint <- stepmeans[stepmeans$meansteps == max(stepmeans$meansteps),]
plot(stepmeans$interval,stepmeans$meansteps,type='l',xlab='Time Interval (min)',ylab='Average Daily Steps',main='Average Steps Taken in each Time Interval')
```
  
The interval that contains the maximum average number of steps is `r maxstepint$interval` minutes.
  
## Imputing missing values
```{r}
totalmissingvals <- sum(is.na(dforig$steps))
uints <- unique(dforig$interval)
dffilled <- read.csv(dataname)
for (tpts in uints) {
  dffilled[dffilled$interval == tpts & is.na(dffilled$steps),1] = round(stepmeans[stepmeans$interval == tpts,2])
}
totalmissingvals1 <- sum(is.na(dffilled$steps))

stepsumsf <- as.data.frame(dffilled %>% group_by(date) %>% summarise(totalsteps=sum(steps), .groups='drop'))
meanstepsf <- mean(stepsumsf$totalsteps)
medstepsf <- median(stepsumsf$totalsteps)
dmeanf <- meanstepsf - meansteps
dmedf <- medstepsf - medsteps
hist(stepsumsf$totalsteps,xlab="Steps Per Day",main="Histogram of Daily Step Count (NAs Filled)")
```
  
The total number of missing values in the dataset is `r totalmissingvals`.  
The missing values were filled with the mean of the interval for all days where data was available.  
The mean total number of daily steps in the new dataset is `r meanstepsf`.  The median is `r medstepsf`.  
The difference between the mean with imputed values and the mean with missing values is `r dmeanf`.  The difference between medians is `r dmedf`.  

## Are there differences in activity patterns between weekdays and weekends?
```{r}
library(lattice)
dffilled$date <- as.Date(dffilled$date)
dffilled <- dffilled %>% mutate(daytype = factor(weekdays(date) == "Saturday" | weekdays(date) == "Sunday", labels = c("weekday","weekend")))
avgww <- as.data.frame(dffilled %>% group_by(interval,daytype) %>% summarise(meansteps=mean(steps), .groups='drop'))
avgww <- avgww[order(avgww$daytype,avgww$interval),]
xyplot(meansteps~interval | daytype, data = avgww,
       type='l',
       layout=c(1,2),
       xlab="Interval (minutes)",
       ylab="Average Steps")
```
  
On weekdays, there is a high peak in steps at `r maxstepint$interval` minutes and on weekends activity is more evenly distributed throughout the day.  
